A graph-based data model for interactive evolutionary computation

### 2017-01-20 2017-CEC-Graph-Based-Population-For- IEC/paper.tex
Spell check

\section{Introduction}
Interactive Evolutionary Algorithms (IEAs) are standard EAs whose
fitness evaluations are performed by persons within an interactive 
system.  Thus, the main loop of the EA needs the intervention of a
human for the quality assessment of each proposed solution.
IEAs have demonstrated their ability for effectively
producing art and design \cite{Bentley:1999:intro,Sims:1991,todd:1992},
as well as other products in many other domains \cite{ie1}. 
Several IECs have been developed as web based systems \cite{picbreeder},
who depend on volunteers who visit the system to interactively help
with the search, using both anonymous and registered users.
These systems have several inherent drawbacks arising from the very nature of 
the algorithms, namely, the human fatigue caused by the interaction, and
the boredom arising when users evaluate a large number of individuals 
many of which are not interesting or are very similar to each other.
IEAs also share some of the issues commonly found in volunteer based systems
\cite{sarmenta2001volunteer,web:BOINC} such as the volunteer\'s lack of accountability,
and the need to build trust between participants and application providers. 
For the provider there is also the difficulty of establishing 
the amount of time and resources
a volunteer is willing to spend on the system, and how they decide if they
participate or not \cite{JJ:2016}. In order to increase participation and 
engagement, and to tackle the issues mentioned above,  
IEAs must follow a user centered design, giving extensive attention 
to volunteer users, not only because their
explicit evaluation is essential, but also because the context of the 
interaction affects the system as a whole. Users interact not only with the graphical
interface, and each individual\'s phenotype, there is also the interaction
with other users, their previous evaluations and their own stream of activities
and past experiences. Thus users in this proposal are defined as dynamic entities 
interacting with an evolutionary computation having the fundamental purpose 
of evaluating individuals according to their preferences,
both explicitly or implicitly. In this work we propose giving the same
importance to users and their interactions as the population of 
individuals have in a traditional EA, even to the extend as using this
knowledge as an integral part of the evolutionary algorithm. We argue also
that a graph model is a viable representation for this purpose,  
presenting a natural way of maping the relationships found in the system. 
Most IEAs have the following many-to-many relations:

@@

  \item {\bf User-User}
  Relationship between users can also be modeled 
  as a social network, with well stablished semantics, algorithms and metrics (cite).
  A graph model could eneable researchers to find other ways of identifyng leaders of 
  opinion or measuring the similarity between user's preferences. 
  These measures can then be used by recommender algorithms selecting 
  individuals according users' preferences.

@@

These relations are naturaly maped to graphs, while other actions and 
properties can also be modeled as a semantic network using
basic subject–-predicate-–object expressions [cite Semantic triples].
Another reason for using a graph model is the tools available for
implementing and query the model, and the increased performance of
native graph database systems [cite Christian].

@@

A case study is presented as a proof of concept, providing both
conceptual and implementation details of a graph data model applied in the 
implementation of an IEA. The experiment was implmented using
the EvoSpace-Interactive framework, and consisted of a web application
for evolving artistic drawings. The graph database is then used to
measure the degree of relationship between users and their participation.

@@

\section{EvoSpace-Interactive}
\label{sec:evospace-i} 
As a case study, a IEA was implented using the 
EvoSpace-Interactive (ES-I) platform \cite{garcia2013evospace}. %citation
A brief description of the aplication is presented next, focusing
on the data elements that where ported to the graph model.

@@

 \section{IEC Human-centered framework}
\label{sec:graph}
The general goal of this research is to develop a human-centered [cite] framework
for interactive evolutionary computation (IEC) in order to increase
participation and also to minimize the amount of evaluations needed for the
evolutionary process in given IEC application. We deliberately use the 
more general
term Human instead of User, because humans are not always playing the role of
users in software engineering sense, they could just be interacting 
naturaly with their environment and with out knowing they are implicitly
evaluating a possible solution. In other cases the evaluation could be done
by the natural interactions of a group of humans in a collaborative fashion
or not. For instance the IEC algorithm could by trying to optimize 
the design of an exit sign in an airport, then the quality of this sign could be
determined by the average the time it takes a sample of people to 
find the exit. In this example there could be certain collaboration
between users if they are not traveling alone, of course this is only one of
the considerations designers could make.

@@

\begin{itemize}
  \item {\bf Interactive Evolutionary Computational system} 
  This is the real world system that we are going to represent in the model, 
  it consist of human users and their interactions with one or more individuals
  from
  the population. There are many ways in which humans could interact 
  with indivuals, this could be through a mobile device, a user interface 
  or by interacting with real world objects generated as part of the algorithm 
  [cite evodraw o unplugged].
  There is also the posibility that fitness or even part of the search 
  to be done by devices lended by the human [cite nodio].
  The information gathered through the interaction is the primary concern
  as it will guide the search.

  \item {\bf IEC Data Model}
  A data model is used to describe the IEC system prior to a physical 
  implementation. 
  %% This block is maybe not needed but in our ES-I implementation
  %% we use Relational (PosgreSQL), Objects (JSON-Redis) and Graphs (Neo4J)
  Depending on the domain several techniques can be used.
  Parts of the system could be better described by an Entity relationship 
  modeling to be used in a relational database or a Class Digram for a 
  key-value store. 
  %%
  %%
  A graph is proposed for modelling the social network
  of users and their interaction with the population. This database 
  is the back-end of the system, is used by the EA and the user interface.

  \item {\bf Evolutionary Algorithm} 
  The EA algorithm queries the graph model, and then updates those nodes that 
  are part of the population. The EA could also be done with the help of
  humans, for instance in the XY project [cite] artists physicaly created
  the new population.  
\end{itemize}

@@

\item {\bf Users} This nodes represent the human users of the system and have 
the attributes: {\tt id} a unique identifier, {\tt name}  and {\tt created}
specifing the  creation date (time stamp) of the node.

@@

\item {\bf Knows} The relation connects two users that know each
other. This relation is "unique" meaning that it can not be repeted between a pair of nodes.

@@

Is important to notice that this model is still partial view of the IEC system,
and other models can could be used in practice when implementing the system. 
For instance a relational database can be used for the implementation of a
user profile and authentication module. Also a copy or cache of the population 
could be kept in a key-value store in memmory for performance reasons. 
Nevertheless, the model can give important insights of the behaviour of the 
system.

@@

We present in this section the results of a comparassion between two versions
of the same IEC application. The goal of the experiment is to compare two 
gamification techiniques to increase volunteer participation. 
The main difference between both versions is how the fitness of individuals
is calculated and the information presented to users. 
\subsection{Gamification technique}
\label{sec:gamification}
The gammification technique
consists in giving more importance to the preference of users with more
participations. A feature found in video games is the concept of a 
score and experience levels. Levels depend on the score and certain features
of the game are only available to gamers when they reach a giving level.

@@

The comparation consisted on three versions:
\begin{itemize}
\item Non Gammification. All users have the same weight.
\item Non Graph Gammification. Only experience is considered
\item Graph Gammification. Both experience and participation where considered.
\end{itemize}
When gammification was employed, all score values known where presented to users
and a ranking of users by weight was shown in a window, a screen capture of
this interface is shown in figure \ref{fig:top-users}.

>>>>

\section{Introduction}
Interactive Evolutionary Algorithms (IEAs) are standard EAs whose
fitness evaluations are performed by persons within an interactive 
system.  Thus, the main loop of the EA needs the intervention of a
human to perform the quality assessment of each proposed solution.
IEAs have demonstrated their ability for effectively
producing art and design \cite{Bentley:1999:intro,Sims:1991,todd:1992},
as well as other products in many other domains \cite{ie1}. 
Several IECs have been developed as web based systems \cite{picbreeder},
who depend on volunteers who visit the system to interactively help
with the search, using both anonymous and registered users.
These systems have several inherent drawbacks arising from the very nature of 
the algorithms, namely, the human fatigue caused by the interaction, and
the boredom arising when users evaluate a large number of individuals 
many of which are not interesting or are very similar to each other.
IEAs also share some of the issues commonly found in volunteer based systems
\cite{sarmenta2001volunteer,web:BOINC} such as the volunteer\'s lack of accountability,
and the need to build trust between participants and application providers. 
For the provider there is also the difficulty of establishing 
the amount of time and resources
a volunteer is willing to spend on the system, and how they decide if they
participate or not \cite{JJ:2016}. In order to increase participation and 
engagement, and to tackle the issues mentioned above,  
IEAs must follow a human centered design, giving extensive attention 
to volunteer users, not only because their
explicit evaluation is essential, but also because the context of the 
interaction affects the system as a whole. Users interact not only with the graphical
interface, and each individual\'s phenotype, there is also the interaction
with other users, their previous evaluations and their own stream of activities
and past experiences. Thus users in this proposal are defined as dynamic entities 
interacting with an evolutionary computation having the fundamental purpose 
of evaluating individuals according to their preferences,
both explicitly or implicitly. In this work we propose giving the same
importance to users and their interactions as the population of 
individuals have in a traditional EA, even to the extend as using this
knowledge as an integral part of the evolutionary algorithm. We argue also
that a graph model is a viable representation for this purpose,  
presenting a natural way of mapping the relationships found in the system. 
Most IEAs have the following many-to-many relations:

@@

  \item {\bf User-User}
  Relationship between users can also be modeled 
  as a social network, with well established semantics, algorithms and metrics (cite).
  A graph model could enable researchers to find other ways of identifying leaders of 
  opinion or measuring the similarity between user's preferences. 
  These measures can then be used by recommender algorithms selecting 
  individuals according users' preferences.

@@

These relations are naturally mapped to graphs, while other actions and 
properties can also be modeled as a semantic network using
basic subject–-predicate-–object expressions [cite Semantic triples].
Another reason for using a graph model is the tools available for
implementing and query the model, and the increased performance of
native graph database systems [cite Christian].

@@

A case study is presented as a proof of concept, providing both
conceptual and implementation details of a graph data model applied in the 
implementation of an IEA. The experiment was implemented using
the EvoSpace-Interactive framework, and consisted of a web application
for evolving artistic drawings. The graph database is then used to
measure the degree of relationship between users and their participation.

@@

\section{EvoSpace-Interactive}
\label{sec:evospace-i} 
As a case study, a IEA was developed using the 
EvoSpace-Interactive (ES-I) platform \cite{garcia2013evospace}. %citation
A brief description of the application is presented next, focusing
on the data elements that where ported to the graph model.

@@

 \section{IEC Human-centered framework}
\label{sec:graph}
The general goal of this research is to develop a human-centered [cite] framework
for interactive evolutionary computation (IEC) in order to increase
participation and also to minimize the amount of evaluations needed for the
evolutionary process in given IEC application. We deliberately use the 
more general
term Human instead of User, because humans are not always playing the role of
users in software engineering sense, they could just be interacting 
naturally with their environment and with out knowing they are implicitly
evaluating a possible solution. In other cases the evaluation could be done
by the natural interactions of a group of humans in a collaborative fashion
or not. For instance the IEC algorithm could by trying to optimize 
the design of an exit sign in an airport, then the quality of this sign could be
determined by the average the time it takes a sample of people to 
find the exit. In this example there could be certain collaboration
between users if they are not traveling alone, of course this is only one of
the considerations designers could make.

@@

\begin{itemize}
  \item {\bf Interactive Evolutionary Computational system} 
  This is the real world system that we are going to represent in the model, 
  it consist of human users and their interactions with one or more individuals
  from
  the population. There are many ways in which humans could interact 
  with individuals, this could be through a mobile device, a user interface 
  or by interacting with real world objects generated as part of the algorithm 
  [cite evodraw o unplugged].
  There is also the possibility that fitness or even part of the search 
  to be done by devices lent by the human [cite nodio].
  The information gathered through the interaction is the primary concern
  as it will guide the search.

  \item {\bf IEC Data Model}
  A data model is used to describe the IEC system prior to a physical 
  implementation. 
  %% This block is maybe not needed but in our ES-I implementation
  %% we use Relational (PosgreSQL), Objects (JSON-Redis) and Graphs (Neo4J)
  Depending on the domain several techniques can be used.
  Parts of the system could be better described by an Entity relationship 
  modeling to be used in a relational database or a Class Digram for a 
  key-value store. 
  %%
  %%
  A graph is proposed for modeling the social network
  of users and their interaction with the population. This database 
  is the back-end of the system, is used by the EA and the user interface.

  \item {\bf Evolutionary Algorithm} 
  The EA algorithm queries the graph model, and then updates those nodes that 
  are part of the population. The EA could also be done with the help of
  humans, for instance in the XY project [cite] artists physically created
  the new population.  
\end{itemize}

@@

\item {\bf Users} This nodes represent the human users of the system and have 
the attributes: {\tt id} a unique identifier, {\tt name}  and {\tt created}
specifying the  creation date (time stamp) of the node.

@@

\item {\bf Knows} The relation connects two users that know each
other. This relation is "unique" meaning that it can not be repeated between a pair of nodes.

@@

Is important to notice that this model is still partial view of the IEC system,
and other models can could be used in practice when implementing the system. 
For instance a relational database can be used for the implementation of a
user profile and authentication module. Also a copy or cache of the population 
could be kept in a key-value store in memory for performance reasons. 
Nevertheless, the model can give important insights of the behavior of the 
system.

@@

We present in this section the results of a comparison between two versions
of the same IEC application. The goal of the experiment is to compare two 
gamification techniques to increase volunteer participation. 
The main difference between both versions is how the fitness of individuals
is calculated and the information presented to users. 
\subsection{Gamification technique}
\label{sec:gamification}
The gamification technique
consists in giving more importance to the preference of users with more
participations. A feature found in video games is the concept of a 
score and experience levels. Levels depend on the score and certain features
of the game are only available to gamers when they reach a giving level.

@@

The comparation consisted on three versions:
\begin{itemize}
\item Non Gamification. All users have the same weight.
\item Non Graph Gamification. Only experience is considered
\item Graph Gamification. Both experience and participation where considered.
\end{itemize}
When gamification was employed, all score values known where presented to users
and a ranking of users by weight was shown in a window, a screen capture of
this interface is shown in figure \ref{fig:top-users}.
